```{r, echo=FALSE}
Sys.setenv(LANG = "en")
SOLS=TRUE
knitr::opts_chunk$set(curq = 0)
newq <- function() {
  knitr::opts_chunk$set(curq = knitr::opts_chunk$get("curq")+1)
  return(paste(knitr::opts_chunk$get("curq"),".",sep=""))
}
getq <- function() {
  return(paste(knitr::opts_chunk$get("curq"),".",sep=""))
}
knitr::opts_chunk$set(curex = 1)
newex <- function() {
  knitr::opts_chunk$set(curex = knitr::opts_chunk$get("curex")+1)
  return(paste(knitr::opts_chunk$get("curex"),"",sep=""))
}
getex <- function() {
  return(paste(knitr::opts_chunk$get("curex"),"",sep=""))
}
```

---
title: "Introduction to Process Mining"
author: "Frédéric Bertrand and Yoann Valero"
date: "08/09/2022"
output: 
  html_document:
    fig_height: 6
    fig_width: 12
  pdf_document:
    fig_height: 6
    fig_width: 12
  word_document:
    fig_height: 6
    fig_width: 12
---

```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

### Setup

You need to install several `R` packages from the BupaRverse.
```{r, eval=FALSE}
install.packages("bupaR")
install.packages("edeaR")
install.packages("eventdataR")
install.packages("processmapR")
install.packages("processanimateR")
install.packages("petrinetR")
```

<div class="alert alert-info">
### Slide 8
</div>


```{r}
library(bupaR)
```

bupaR autmaticaly loads eventdataR

```{r}
#library(eventdataR)
```

#### Retrieve example datasets from the eventdataR

When loading the bupaR package, the `traffic_fines` dataset is also loaded into memory.

#### Print the case long format table (slide 8)
```{r}
traffic_fines
```

We have to turn the `traffic_fines` dataset into an eventlog/log class using the `simple_eventlog` function.

```{r}
log_traffic_fines <- simple_eventlog(traffic_fines)
```

For this dataset, you can also just add the "log" class to the `R` object.

```{r, eval=FALSE}
log_traffic_fines <- traffic_fines
class(log_traffic_fines) <- c("log",class(log_traffic_fines))
```



<div class="alert alert-success">
`r newq()` **Hands-on training** From the same `eventdataR` package, retrieve the datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>


<div class="alert alert-info">
### Slide 9
</div>

#### Print the case long format table (slide 9)

```{r}
case_list(log_traffic_fines)
```

A more detailed view on lists:
```{r}
cases(log_traffic_fines)
```

#### Rapid glance to the cases' length distribution
```{r}
case_length_dist_tf <- sapply(strsplit(case_list(log_traffic_fines)$trace, ","), length) 
```


```{r}
table(case_length_dist_tf)
```

```{r}
barplot(table(case_length_dist_tf))
```

```{r}
quantile(case_length_dist_tf,probs = seq(0, 1, 0.1))
```


```{r}
log_traffic_fines %>%
  durations(units="hours") %>% # duration of the cases
  as.data.frame(units="hours") %>% # cases and their duration
  ggplot(aes(x=duration))+
  geom_density(color="#1F77b4",
                 fill="#17BECF",
                 alpha=.5)+
  theme_minimal()
```



#### Print the trace table (slide 9)

```{r}
trace_list(log_traffic_fines)
```

<div class="alert alert-success">
`r newq()` **Hands-on training** Display the case and trace tables as well as the cases' lengths distribution for the following datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>


For instance, here are the results for the sepsis log.
```{r, echo=FALSE, cache=TRUE}
sepsis
log_sepsis <- simple_eventlog(sepsis)
case_list(log_sepsis)
cases(log_sepsis)
case_length_dist_se <- sapply(strsplit(case_list(log_sepsis)$trace, ","), length) 
table(case_length_dist_se)
barplot(table(case_length_dist_se))
quantile(case_length_dist_se,probs = seq(0, 1, 0.1))
log_sepsis %>%
  durations(units="hours") %>% # duration of the cases
  as.data.frame(units="hours") %>% # cases and their duration
  ggplot(aes(x=duration))+
  geom_density(color="#1F77b4",
                 fill="#17BECF",
                 alpha=.5)+
  theme_minimal()
```


<div class="alert alert-info">
### Slide 10
</div>

The `print` function displays a summary of the log. The same result is provided by just typinf the log's object name in the `R`console.
```{r}
print(log_traffic_fines)
```

To get the number of events:
```{r}
n_events(log_traffic_fines)
```

To get the number of cases:
```{r}
n_cases(log_traffic_fines)
```

To get the number of traces:
```{r}
n_traces(log_traffic_fines)
```

To get the number of activities:
```{r}
n_activities(log_traffic_fines)
```

Recall the formula for *Average case length*:

$$\displaystyle\frac{Event}{Cases}=\frac{\mathrm{n\_events(log\_traffic\_fines)}}{\mathrm{n\_cases(log\_traffic\_fines)}}$$

```{r}
n_events(log_traffic_fines)/n_cases(log_traffic_fines)
```


$\displaystyle\frac{Event}{Cases}=`r n_events(log_traffic_fines)/n_cases(log_traffic_fines)`$

Recall the formula for *Process uniformity*:

$$\displaystyle 1−\frac{Traces-1}{Cases}=1-\frac{\mathrm{n\_traces(log\_traffic\_fines)-1}}{\mathrm{n\_cases(log\_traffic\_fines)}}$$

```{r}
1-(n_traces(log_traffic_fines)-1)/n_cases(log_traffic_fines)
```


$\displaystyle\frac{Event}{Cases}=`r 1-(n_traces(log_traffic_fines)-1)/n_cases(log_traffic_fines)`$


<div class="alert alert-success">
`r newq()` **Hands-on training** Compute the average case length and the process uniformity for the following datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>




<div class="alert alert-info">
### Slide 13
</div>


Compute and display the *Precedence Matrix* for absolute frequencies.

```{r, cache=TRUE}
process_matrix(log_traffic_fines)
```

```{r, cache=TRUE}
plot(process_matrix(log_traffic_fines))
```

It is a ggplot object that can be customized.

```{r, cache=TRUE}
plot(process_matrix(log_traffic_fines))+
  geom_tile(aes(fill=n))+
  geom_text(aes(label=n),color="white",size=2.5)+
  ggtitle("Traffic fines process map")+
  scale_fill_viridis_c(option="turbo")+
  theme_light()+
  theme(axis.text.x=element_text(angle=45,
                                 hjust=1,
                                 size=8),
        axis.text.y=element_text(size=8))
```

In addition to the absolute frequencies, other quantities can be computed:

* "absolute", 
* "relative", 
* "absolute-case", 
* "relative-case", 
* "relative-antecedent", 
* "relative-consequent",

where absolute, relative (percentage of activity instances) or relative_case (percentage of cases the activity occurs in).


```{r, cache=TRUE}
plot(process_matrix(log_traffic_fines,type=frequency(value="relative")))+geom_tile(aes(fill = rel_n))+geom_text(aes(label=round(rel_n * 100, 2)),color="white",size=2.5)
```

```{r, cache=TRUE}
plot(process_matrix(log_traffic_fines,type=frequency(value="absolute-case")))+geom_tile(aes(fill = n_cases))+geom_text(aes(label=n_cases),color="white",size=2.5)
```

<div class="alert alert-success">
`r newq()` **Hands-on training** Compute and display precedence matrices for the following datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>



<div class="alert alert-info">
### Slide 17
</div>

The percentage coverage of the trace to explore. Default is 20% most (in)frequent
```{r, cache=TRUE}
trace_explorer(log_traffic_fines, coverage=1)
```

Instead of setting coverage, you can set an exact number of traces. Should be an integer larger than 0.
```{r, cache=TRUE}
trace_explorer(log_traffic_fines, n_traces = 10)
```

```{r}
library(TraMineR)
```

```{r echo=FALSE, message=FALSE}
cl_tr<-case_list(log_traffic_fines)$trace
s_tf<-cl_tr %>%
  seqdecomp(sep=",") %>%
  seqdef
```


Plot selected cases.
```{r, cache=TRUE}
seqplot(s_tf,
        type="i",  
        cex.axis=.8,
        cex.legend=.8,
        legend.prop = .44, 
        with.legend = "right",
)
```

Other types : 

* d, for state distribution plots
* dH, for chronograms with overlayed entropy line
* f, for sequence frequency plots
* i, for selected sequence index plots
* I, for whole set index plots
* Ht, for transversal entropy plots
* ms, for plotting the sequence of modal states
* mt,  for mean times plots
* pc, for parallel coordinate plots


```{r, cache=TRUE}
seqplot(s_tf,
        type="dH",  
        cex.axis=.8,
        cex.legend=.8,legend.prop = .44, with.legend = "right",
)
```

```{r, cache=TRUE}
seqplot(s_tf,
        type="Ht",  
        cex.axis=.8,
        cex.legend=.8,
        legend.prop = .44, 
        with.legend = "right",
)
```

```{r, cache=TRUE}
seqplot(s_tf,
        type="pc",  
        cex.axis=.8,
        cex.legend=.8,
        legend.prop = .44, 
        with.legend = "right",
)
```

To get the state frequencies use, the `seqstatd` function.
```{r}
seqstatd(s_tf, weighted=FALSE)
```

<div class="alert alert-success">
`r newq()` **Hands-on training** Use the trace_explorer and compute entropies for the following datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>





<div class="alert alert-info">
### Slide 21
</div>

```{r}
library(heuristicsmineR)
process_map(L_heur_1)
```

You can easily filter the data to map using the `%>%` operator.
```{r}
pm<-L_heur_1 %>%
  filter_activity_frequency(percentage=1) %>%
  filter_trace_frequency(percentage=1) %>%
  process_map(sec_nodes=frequency(value="absolute"),
              type=performance(FUN=mean,
                               units="hours",
                               color_scale="Greens",
                               color_edges="Greens"),
              unit="hours")
pm
```

```{r}
pm<-L_heur_1 %>%
  filter_activity_frequency(percentage=.5) %>%
  filter_trace_frequency(percentage=.5) %>%
  process_map(sec_nodes=frequency(value="absolute"),
              type=performance(FUN=mean,
                               units="hours",
                               color_scale="Greens",
                               color_edges="Greens"),
              unit="hours")
pm
```

<div class="alert alert-success">
`r newq()` **Hands-on training** Create those maps for the following datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>


<div class="alert alert-info">
### Slide 25
</div>

Full dataset
```{r}
process_map(log_traffic_fines, rankdir = "TB")
```

Filtering to get a 90% coverage of activity instances. 
```{r}
log_traffic_fines %>%
  filter_trace_frequency(percentage=.9) %>%
process_map(rankdir = "TB")
```

```{r}
log_traffic_fines %>%
  filter_trace_frequency(percentage=.9) %>%
process_map(rankdir = "TB", performance(FUN=mean,
                               units="hours",
                               color_scale="Greens",
                               color_edges="Greens"))
```

<div class="alert alert-success">
`r newq()` **Hands-on training** Create those maps for the following datasets:

a. `sepsis`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>


<div class="alert alert-info">
### Slide 36
</div>

```{r, cache=TRUE}
library("viridisLite")
m<-log_sepsis %>%
  dependency_matrix(threshold=0) %>%
  (reshape2::melt)
m[m$value!=0,] %>%
  ggplot(aes(consequent, antecedent, label=round(value,2)))+
  geom_tile(aes(fill=value), color="grey27")+
  geom_text(color="white",size=2.5)+
  theme_light()+
  theme(axis.text.x=element_text(angle=45,
                                 hjust=1,
                                 size=8),
        axis.text.y=element_text(size=8))
```

```{r, cache=TRUE}
library("viridisLite")
m<-log_sepsis %>%
  dependency_matrix(threshold=.9) %>%
  (reshape2::melt)
m[m$value!=0,] %>%
  ggplot(aes(consequent, antecedent, label=round(value,2)))+
  geom_tile(aes(fill=value), color="grey27")+
  geom_text(color="white",size=2.5)+
  theme_light()+
  theme(axis.text.x=element_text(angle=45,
                                 hjust=1,
                                 size=8),
        axis.text.y=element_text(size=8))
```

<div class="alert alert-info">
### Slide 39
</div>

To compute the process map, first:

```{r}
sepsis
log_sepsis <- simple_eventlog(sepsis)
```

```{r, cache=TRUE}
process_map(log_sepsis, rankdir = "TB")
```

Quite difficult to use, we need to mine it.

<div class="alert alert-info">
### Slide 40
</div>

To compute the causal net, first:

#### Dependency matrix 


```{r}
sepsis
log_sepsis <- simple_eventlog(sepsis)
```

```{r, cache=TRUE}
library(heuristicsmineR)
plot(dependency_matrix(log_sepsis))+geom_raster(aes(fill = dep))+geom_text(aes(label=round(dep, 2)),color="white",size=2.5)
```

The option `dependency_type` enables to tune the computation of the dependency matrix. For instance, `dependency_type=dependency_type_fhm(...)` where `...` stands for:

```
dependency_type_fhm(
  threshold_dependency = 0.9,
  threshold_l1 = threshold_dependency,
  threshold_l2 = threshold_dependency,
  threshold_frequency = 0,
  all_connected = FALSE,
  endpoints_connected = FALSE
)
```

with 


* `threshold_dependency` A dependency threshold, usually in the interval [0,1], filtering out dependencies below the threshold.

* `threshold_l1` A dependency threshold, usually in the interval [0,1], filtering out self-loop dependencies below the threshold.

* `threshold_l2` A dependency threshold, usually in the interval [0,1], filtering out length-two loop dependencies below the threshold.

* `threshold_frequency` An absolute frequency threshold filtering dependencies which are observed infrequently.

* `all_connected` If TRUE the best antecedent and consequent (as determined by the dependency measure) are going to be added regardless of the threshold value.

* `endpoints_connected` If TRUE the start/end activity is added as antecedent/consequent when an activity would not be connected according to the threshold value.

```{r, cache=TRUE}
library("viridisLite")
m<-log_sepsis %>%
  dependency_matrix(threshold=.8) %>%
  (reshape2::melt)
m[m$value!=0,] %>%
  ggplot(aes(consequent, antecedent, label=round(value,2)))+
  geom_tile(aes(fill=value), color="grey27")+
  scale_fill_gradientn(colors=viridis(256),
                       name="Dependency\nmeasure")+
  theme_light()+
  theme(axis.text.x=element_text(angle=45,
                                 hjust=1,
                                 size=8),
        axis.text.y=element_text(size=8))
```

```{r, cache=TRUE}
cn_sep0<-causal_net(log_sepsis)
cn_sep0
```


*Consider using the `all_connected` or `endpoints_connected` parameter when generating the dependency matrix.*

```{r, cache=TRUE}
cn_sep<-causal_net(log_sepsis, all_connected = TRUE)
cn_sep
```


<div class="alert alert-success">
`r newq()` **Hands-on training** Compute precedence matrices and causal nets for the following datasets:

a. `traffic_fines`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>


<div class="alert alert-info">
### Slide 40
</div>


```{r}
pn1<-(as.petrinet(cn_sep))
unique(pn1$flows$to[!pn1$flows$to %in% pn1$flows$from])
render_PN(pn1)
```

Place/Transition Connecting Degree (PTCD)
```{r}
(n_flows(pn1)/n_places(pn1)+n_flows(pn1)/n_transitions(pn1))/2
```



<div class="alert alert-success">
`r newq()` **Hands-on training** Compute and display precedence matrices for the following datasets:

a. `traffic_fines`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>


<div class="alert alert-info">
### Animate process maps
</div>


```{r}
library(processanimateR)
animate_process(log_traffic_fines)
```


A more intricate example, from 
```{r}
library(processanimateR)
library(dplyr)
library(bupaR)
# Extract only the lacticacid measurements
lactic <- log_sepsis %>%
    mutate(lacticacid = as.numeric(lacticacid)) %>%
    filter_activity(c("LacticAcid")) %>%
    as.data.frame() %>%
    select("case" = case_id, 
            "time" =  timestamp, 
            value = lacticacid) # format needs to be 'case,time,value'

# Remove the measurement events from the sepsis log
sepsisBase <- log_sepsis %>%
    filter_activity(c("LacticAcid", "CRP", "Leucocytes", "Return ER",
                      "IV Liquid", "IV Antibiotics"), reverse = T) %>%
    filter_trace_frequency(percentage = 0.95)

# Animate with the secondary data frame `lactic`
animate_process(sepsisBase, 
                mode = "relative", 
                duration = 300,
                legend = "color", 
                mapping = token_aes(color = token_scale(lactic, 
                                                        scale = "linear", 
                                                        range = c("#fff5eb","#7f2704")))) 
```

<div class="alert alert-success">
`r newq()` **Hands-on training** Display animate the processes of the following logs:

a. `traffic_fines`
b. `hospital_billing`
c. `hospital`
d. `patients`
</div>